Notes on filtration of logics containing {\bf K5}

### 2021-03-04 filtration-stuff/Text.tex
fix the proof of Theorem 3

Let $\mathcal{L}$ be a normal modal logic, $\mathcal{M}_{\mathcal{L}}$ its canonical model, and $\varphi$. Let us put $\Gamma = \operatorname{Sub}(\varphi) \cup \{ \neg \psi \: | \: \psi \in \operatorname{Sub}(\varphi) \}$.

A subset $\Delta \subseteq \Gamma$ is a \emph{finite $\mathcal{L}$-consistent set} if $\neg \bigwedge \Delta \notin \mathcal{L}$. A subset $\Delta$ is maximal, if (the following are obviously equivalent):
\begin{enumerate}
  \item $\Delta$ is maximal amongst finite $\mathcal{L}$-consistent sets,
  \item For each $\psi \in \operatorname{Sub}(\varphi)$ either $\psi \in \Delta$ or
  $\neg \psi \in \Delta$.
\end{enumerate}

@@

\begin{defin}\label{fin}
  Let $\mathcal{L}$ be a normal modal logic and $\varphi \not\in \mathcal{L}$. A finite ``canonical'' model is a triple $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$, where
  \begin{enumerate}
    \item $W_{\mathcal{L}}^{\varphi}$ is the set all maximal theories that extend finite $\mathcal{L}$-theories
    \item $R_{\mathcal{L}}^{\varphi}$ is a relation such that
    $\langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi} \rangle$ is an $\mathcal{L}$-frame
    and
    \begin{center}
    $\forall \Box \psi \in \operatorname{Sub}(\varphi) \:\:
    \forall \Delta_1 \in W_{\mathcal{L}}^{\varphi} \:\: (\Box \psi \in \Delta_1 \Leftrightarrow
    \forall \Delta_2 \in R_{\mathcal{L}}^{\varphi}(\Delta_1) \:\: \psi \in \Delta_2)$
    \end{center}
    \item $\vartheta_{\mathcal{L}}^{\varphi}(p) = \{ \Delta \in W_{\mathcal{L}}^{\varphi} \: | \: p \in \Delta \}$.
  \end{enumerate}
\end{defin}

\begin{theorem} $\mathcal{L}$ be a modal logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\varphi} \models \mathcal{L}$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$.
  \end{enumerate}
\end{theorem}

@@

  \begin{enumerate}
    \item $(\Rightarrow)$
    Let $\varphi \not\in \mathcal{L}$, let us put $\Gamma = \operatorname{Sub}(\varphi)$.
    $\mathcal{L}$.
    In particular, $\mathcal{M}_{\mathcal{L}} \nvDash \varphi$, where $\mathcal{M}_{\mathcal{L}}$ is the canonical model of $\mathcal{L}$. $\mathcal{L}$ admits strict filtrations, so the filtration of $\mathcal{M}_{\mathcal{L}}$ through $\Gamma$ is also an $\mathcal{L}$-model.
    The underlying set of $\mathcal{M}_{\mathcal{L}} / \sim_{\Gamma}$ consists of maximal $\mathcal{L}$ theories up to $\Gamma$-equivalence and this quotient set is finite.

@@

    Let $\varphi \not\in \mathcal{L}$ and $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$
    be a finite ``canonical'' model that refutes $\varphi$.
    Let $\mathcal{M}$ be an $\mathcal{L}$-model that also refutes $\varphi$.
    Let us show that there exists a model $\widehat{\mathcal{M}}$ such that
    $\widehat{\mathcal{M}}$ is a strict filtration of $\mathcal{M}$ through $\operatorname{Sub}(\varphi)$.

    Let $\sim_{\operatorname{Sub}(\varphi)}$ be an equivalence relation on $\underline{\mathcal{M}}$ defined as usual:

    \begin{center}
      $u \sim_{\operatorname{Sub}(\varphi)} v$ iff $\forall \psi \in \operatorname{Sub}(\varphi)$
      $\mathcal{M}, u \models \psi \Leftrightarrow \mathcal{M}, v \models \psi$
    \end{center}

    Let $[u] \in W/\sim_{\operatorname{Sub}(\varphi)}$, where $W = \underline{\mathcal{M}}$,
    consider its truth set $||[u]|| = \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite $\mathcal{L}$-consistent set and it is maximal.
    Consider a map
    \begin{center}
    $\iota : W/\sim_{\operatorname{Sub}(\varphi)} \to W_{\mathcal{L}}^{\varphi}$ with
    $\iota : [u] \mapsto ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim_{\operatorname{Sub}(\varphi)})$ that we denote as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\varphi}$-closed.

    Let $\iota([u]) R_{\mathcal{L}}^{\varphi} \Delta$, let us show that $\Delta \in \iota(W/\sim_{\operatorname{Sub}(\varphi)})$, that is, $\Delta = \iota([v])$ for some $v \in W$.

    Note that $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta$ since $\Delta$ is finite.
    This conjuction has the form $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta = \psi_1 \land \dots \land \psi_n$ for some $n < \omega$ and for some $\psi_i \in \operatorname{Sub}(\varphi) \cup \neg \operatorname{Sub}(\varphi)$, where $0 < i < \omega$.

    Therefore $\mathcal{M}^{\varphi}_{\mathcal{L}}, \iota([u]) \models \Diamond (\psi_1 \land \dots \land \psi_n)$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\varphi}, {\vartheta'}_{\mathcal{L}}^{\varphi} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\varphi}(p) = \vartheta_{\mathcal{L}}^{\varphi}(p) \cap \operatorname{Im}(\iota)$
    is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\varphi}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim_{\operatorname{Sub}(\varphi)}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R}_{\mathcal{L}}^{\varphi} ||[v]||$
    \end{center}

@@

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(v)$.
    But if $\mathcal{M}, v' \models \psi$ for $\Diamond \psi \in \operatorname{Sub}(\varphi)$, then
    $\mathcal{M}, u' \models \Diamond \psi$. Recall that $u' \sim_{\operatorname{Sub}(\varphi)} u$ and $v' \sim_{\operatorname{Sub}(\varphi)} v$, so we have $\Diamond ||[v]|| \subseteq ||[u]||$.

>>>>

Let $\mathcal{L}$ be a normal modal logic, $\mathcal{M}_{\mathcal{L}}$ its canonical model, and $\Gamma$ a finite $\operatorname{Sub}$-closed set of formulas. Let us put $\Gamma' = \operatorname{Sub}(\varphi) \cup \{ \neg \psi \: | \: \psi \in \operatorname{Sub}(\varphi) \}$.

A subset $\Delta \subseteq '\Gamma$ is a \emph{finite $\mathcal{L}$-consistent set} if $\neg \bigwedge \Delta \notin \mathcal{L}$. A subset $\Delta$ is maximal, if (the following are obviously equivalent):
\begin{enumerate}
  \item $\Delta$ is maximal amongst finite $\mathcal{L}$-consistent sets,
  \item For each $\psi \in \operatorname{Sub}(\varphi)$ either $\psi \in \Delta$ or
  $\neg \psi \in \Delta$.
\end{enumerate}

@@

\begin{defin}\label{fin}
  Let $\mathcal{L}$ be a modal logic and $\Gamma$ be a finite $\operatorname{Sub}$-closed set of formulas. A finite ``canonical'' model is a triple $\mathcal{M}_{\mathcal{L}}^{\Gamma} = \langle W_{\mathcal{L}}^{\Gamma}, R_{\mathcal{L}}^{\Gamma}, \vartheta_{\mathcal{L}}^{\Gamma} \rangle$, where
  \begin{enumerate}
    \item $W_{\mathcal{L}}^{\Gamma}$ is the set all maximal theories that extend finite $\mathcal{L}$-theories
    \item $R_{\mathcal{L}}^{\Gamma}$ is a relation such that
    $\langle W_{\mathcal{L}}^{\Gamma}, R_{\mathcal{L}}^{\Gamma} \rangle$ is an $\mathcal{L}$-frame
    and
    \begin{center}
    $\forall \Box \psi \in \operatorname{Sub}(\varphi) \:\:
    \forall \Delta_1 \in W_{\mathcal{L}}^{\Gamma} \:\: (\Box \psi \in \Delta_1 \Leftrightarrow
    \forall \Delta_2 \in R_{\mathcal{L}}^{\Gamma}(\Delta_1) \:\: \psi \in \Delta_2)$
    \end{center}
    \item $\vartheta_{\mathcal{L}}^{\Gamma}(p) = \{ \Delta \in W_{\mathcal{L}}^{\varphi} \: | \: p \in \Delta \}$ for every variable $p \in \Gamma$.
  \end{enumerate}
\end{defin}

\begin{lemma}
  Let $\mathcal{L}$ be a modal logic and $\varphi \not\in \mathcal{L}$, then $\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)} \nvDash \varphi$.
\end{lemma}

\begin{theorem} $\mathcal{L}$ be a modal logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\Gamma$ is a finite $\operatorname{Sub}$-closed set of formulas , there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\Gamma}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\Gamma} \models \mathcal{L}$.
  \end{enumerate}
\end{theorem}

@@

  \begin{enumerate}
    \item $(\Rightarrow)$
  Let $\Gamma$ be a finite $\operatorname{Sub}$-closed of formulas. $\mathcal{L}$ admits strict filtrations, so the filtration of the canonical model $\mathcal{M}_{\mathcal{L}}$ through $\Gamma$ is also an $\mathcal{L}$-model. The underlying set of $\mathcal{M}_{\mathcal{L}} / \sim_{\Gamma}$ consists of maximal $\mathcal{L}$ theories up to $\Gamma$-equivalence and this quotient set is finite.

@@

    Let $\Gamma$ be a finite $\operatorname{Sub}$-closed of formulas and $\mathcal{M}_{\mathcal{L}}^{\Gamma} = \langle W_{\mathcal{L}}^{\Gamma}, R_{\mathcal{L}}^{\Gamma}, \vartheta_{\mathcal{L}}^{\Gamma} \rangle$ a ``finite canonical'' model generated by $\Gamma$.

    Let us show that there exists a model $\widehat{\mathcal{M}}$ such that
    $\widehat{\mathcal{M}}$ is a strict filtration of $\mathcal{M}$ through $\operatorname{Sub}(\Gamma)$.

    Let $\sim$ be an equivalence relation on $\underline{\mathcal{M}}$ defined as usual:

    \begin{center}
      $u \sim v$ iff $\forall \psi \in \Gamma$
      $\mathcal{M}, u \models \psi \Leftrightarrow \mathcal{M}, v \models \psi$
    \end{center}

    Let $[u] \in \mathcal{M}/\sim$, consider its truth set $||[u]|| = \{ \psi \in \Gamma \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite $\mathcal{L}$-consistent set and it is maximal.
    Consider a map
    \begin{center}
    $\iota : W/\sim \to W_{\mathcal{L}}^{\varphi}$ with $\iota : [u] \mapsto ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim)$ that we denote as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\varphi}$-closed.

    Let $\iota([u]) R_{\mathcal{L}}^{\varphi} \Delta$, let us show that $\Delta \in \iota(W/\sim)$, that is, $\Delta = \iota([v])$ for some $v \in W$.

    Note that $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta$ since $\Delta$ is finite.
    This conjuction has the form $\bigwedge \limits_{\psi \in \Delta} \psi \in \Delta = \psi_1 \land \dots \land \psi_n$ for some $n < \omega$ and for some $\psi_i \in \Gamma \cup \neg \Gamma$, where $0 < i < \omega$.

    Therefore $\mathcal{M}^{\Gamma}_{\mathcal{L}}, \iota([u]) \models \Diamond (\psi_1 \land \dots \land \psi_n)$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\Gamma}, {\vartheta'}_{\mathcal{L}}^{\Gamma} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\Gamma}(p) = \vartheta_{\mathcal{L}}^{\Gamma}(p) \cap \operatorname{Im}(\iota)$
    is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\Gamma}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R}_{\mathcal{L}}^{\Gamma} ||[v]||$
    \end{center}

@@

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(v)$.
    But if $\mathcal{M}, v' \models \psi$ for $\Diamond \psi \in \Gamma$, then
    $\mathcal{M}, u' \models \Diamond \psi$. Recall that $u' \sim u$ and $v' \sim v$, so we have $\Diamond ||[v]|| \subseteq ||[u]||$.
